---
title: "bwTools"
format:
  gfm:
    toc: true
execute:
  enabled: false
---

Current development release: **0.4.1**

`bwTools` provides native-R I/O and indexed analysis for BigWig, WIG, and
bedGraph genomic signal tracks. It is designed as the signal backend for
GeneTrackR without requiring UCSC command-line programs, Rcpp, libBigWig, or
other compiled BigWig dependencies.

This `README.qmd` is the source document. Render it with:

```{bash}
quarto render README.qmd
```

Code execution is disabled during rendering. The `{r}` chunks below are intended
to be executed interactively in order when validating a development release.

## Design

- BigWig parsing, writing, zoom construction, and statistics are implemented in R.
- BigWig random access uses chromosome B+ tree and R-tree indexes.
- Full-resolution and zoom blocks use zlib compression through base R.
- Public in-memory coordinates are **1-based closed**.
- BigWig and bedGraph coordinates are **0-based half-open** on disk.
- Large BigWig inputs should normally use `mode = "lazy"`.
- Returned objects inherit from `BwgTrack` for GeneTrackR interoperability.

## Bundled example data

The package includes:

| File | Purpose |
| --- | --- |
| `bwtools_example.bigwig` | BigWig with discontinuous gene-like coverage |
| `bwtools_example.bedGraph` | Plain-text reference signal |
| `bwtools_example.chrom.sizes` | Chromosome lengths |

Chromosome lengths are `chr1 = 10000`, `chr2 = 8000`, and `chr3 = 5000`.

```{r}
library(bwTools)

bw_file <- system.file(
  "extdata",
  "bwtools_example.bigwig",
  package = "bwTools",
  mustWork = TRUE
)

bedgraph_file <- system.file(
  "extdata",
  "bwtools_example.bedGraph",
  package = "bwTools",
  mustWork = TRUE
)

chrom_sizes_file <- system.file(
  "extdata",
  "bwtools_example.chrom.sizes",
  package = "bwTools",
  mustWork = TRUE
)

stopifnot(
  file.exists(bw_file),
  file.exists(bedgraph_file),
  file.exists(chrom_sizes_file)
)
```

For round-trip checks, normalize the signal representation before comparing
files. This keeps the test focused on genomic coordinates and values rather
than row names or storage-specific attributes.

```{r}
canonical_signal <- function(x) {
  columns <- c(
    "sample_id",
    "chrom",
    "start",
    "end",
    "value",
    "strand"
  )

  x <- as.data.frame(x)[, columns, drop = FALSE]
  x$sample_id <- as.character(x$sample_id)
  x$chrom <- as.character(x$chrom)
  x$start <- as.integer(x$start)
  x$end <- as.integer(x$end)
  x$value <- as.numeric(x$value)
  x$strand <- as.character(x$strand)

  x <- x[
    order(
      x$sample_id,
      x$chrom,
      x$start,
      x$end,
      x$strand
    ),
    ,
    drop = FALSE
  ]
  row.names(x) <- NULL
  x
}

assert_signal_equal <- function(expected, observed, tolerance = 1e-4) {
  expected <- canonical_signal(expected)
  observed <- canonical_signal(observed)
  comparison <- all.equal(
    expected,
    observed,
    tolerance = tolerance,
    check.attributes = FALSE
  )

  if (!isTRUE(comparison)) {
    message("Signal round-trip comparison failed:")
    print(comparison)

    if (nrow(expected) != nrow(observed)) {
      message(
        "Row count: expected=",
        nrow(expected),
        ", observed=",
        nrow(observed)
      )
    }

    common_n <- min(nrow(expected), nrow(observed))
    if (common_n > 0L) {
      unequal <- vapply(
        seq_len(common_n),
        function(i) {
          !isTRUE(
            all.equal(
              expected[i, , drop = FALSE],
              observed[i, , drop = FALSE],
              tolerance = tolerance,
              check.attributes = FALSE
            )
          )
        },
        logical(1L)
      )
      first_bad <- which(unequal)[1L]
      if (!is.na(first_bad)) {
        message("First differing row: ", first_bad)
        print(expected[first_bad, , drop = FALSE])
        print(observed[first_bad, , drop = FALSE])
      }
    }

    stop("Signal round-trip validation failed.", call. = FALSE)
  }

  invisible(TRUE)
}
```

## Reader validation

### 1. Detect formats

```{r}
stopifnot(
  identical(bw_detect_format(bw_file), "bigwig"),
  identical(bw_detect_format(bedgraph_file), "bedgraph")
)
```

### 2. Inspect chromosome metadata

```{r}
seqinfo <- seqinfo_bwg(bw_file)
print(seqinfo)

stopifnot(
  identical(seqinfo$chrom, c("chr1", "chr2", "chr3")),
  identical(seqinfo$length, c(10000L, 8000L, 5000L))
)
```

### 3. Read BigWig lazily

```{r}
track_lazy <- read_bwg(
  bw_file,
  format = "bigwig",
  sample_names = "example",
  mode = "lazy"
)

print(track_lazy)
summary_bwg(track_lazy)

stopifnot(
  is_bwg_track(track_lazy),
  is.null(track_lazy$data),
  nrow(track_lazy$samples) == 1L,
  nrow(track_lazy$seqinfo) == 3L
)
```

### 4. Retrieve a region

```{r}
signal_lazy <- retrieve_bwg(
  track_lazy,
  chrom = "chr1",
  start = 450,
  end = 1600
)

print(signal_lazy)

stopifnot(
  nrow(signal_lazy) > 0L,
  all(signal_lazy$chrom == "chr1"),
  all(signal_lazy$start >= 450L),
  all(signal_lazy$end <= 1600L)
)
```

### 5. Compare BigWig with bedGraph

```{r}
track_bedgraph <- read_bwg(
  bedgraph_file,
  format = "bedgraph",
  sample_names = "example",
  mode = "memory"
)

signal_bedgraph <- retrieve_bwg(
  track_bedgraph,
  chrom = "chr1",
  start = 450,
  end = 1600
)

compare_columns <- c(
  "sample_id",
  "chrom",
  "start",
  "end",
  "value",
  "strand"
)

bw_observed <- as.data.frame(signal_lazy)[, compare_columns, drop = FALSE]
bedgraph_observed <- as.data.frame(signal_bedgraph)[, compare_columns, drop = FALSE]

stopifnot(
  isTRUE(
    all.equal(
      bw_observed,
      bedgraph_observed,
      tolerance = 1e-4
    )
  )
)
```

## Writer validation

### 6. Write BigWig natively

```{r}
test_root <- file.path(tempdir(), "bwtools_readme_test")

if (dir.exists(test_root)) {
  unlink(test_root, recursive = TRUE)
}

dir.create(test_root, recursive = TRUE)

written_bigwig <- write_bwg(
  track_bedgraph,
  outdir = test_root,
  format = "bigwig",
  chrom_sizes = chrom_sizes_file,
  overwrite = TRUE
)

print(written_bigwig)

stopifnot(
  nrow(written_bigwig) == 1L,
  file.exists(written_bigwig$file),
  identical(bw_detect_format(written_bigwig$file), "bigwig")
)
```

The native writer creates the BigWig v4 header, chromosome B+ tree,
full-resolution data blocks, full-data R-tree, optional zoom summary blocks,
zoom R-trees, total summary, and terminal magic value entirely in R.

### 7. Validate BigWig round-trip

```{r}
roundtrip_track <- read_bwg(
  written_bigwig$file,
  format = "bigwig",
  sample_names = "example",
  mode = "memory"
)

original_data <- as.data.frame(track_bedgraph$data)
roundtrip_data <- as.data.frame(roundtrip_track$data)

original_data <- original_data[
  order(original_data$chrom, original_data$start, original_data$end),
  ,
  drop = FALSE
]
roundtrip_data <- roundtrip_data[
  order(roundtrip_data$chrom, roundtrip_data$start, roundtrip_data$end),
  ,
  drop = FALSE
]

row.names(original_data) <- NULL
row.names(roundtrip_data) <- NULL

stopifnot(
  isTRUE(
    all.equal(
      original_data,
      roundtrip_data,
      tolerance = 1e-4
    )
  )
)
```

## Zoom-level validation

Do not use the internal `bw_metadata()` structure in user code. Use the public
`zoominfo_bwg()` interface.

### 8. Inspect zoom levels

```{r}
zoom_info <- zoominfo_bwg(written_bigwig$file)
print(zoom_info)

stopifnot(
  nrow(zoom_info) > 0L,
  all(zoom_info$level > 0),
  all(zoom_info$data_offset > 0),
  all(zoom_info$index_offset > zoom_info$data_offset)
)
```

The same information can be obtained from a lazy `BwgTrack`:

```{r}
roundtrip_lazy <- read_bwg(
  written_bigwig$file,
  format = "bigwig",
  sample_names = "example",
  mode = "lazy"
)

zoom_track <- zoominfo_bwg(roundtrip_lazy)
print(zoom_track)

stopifnot(
  nrow(zoom_track) == nrow(zoom_info),
  all(zoom_track$sample_id == "example"),
  identical(zoom_track$level, zoom_info$level)
)
```

### 9. Validate `zoom = FALSE`

```{r}
nozoom_dir <- file.path(test_root, "nozoom")

written_nozoom <- write_bwg(
  track_bedgraph,
  outdir = nozoom_dir,
  format = "bigwig",
  chrom_sizes = chrom_sizes_file,
  overwrite = TRUE,
  zoom = FALSE
)

nozoom_info <- zoominfo_bwg(written_nozoom$file)
print(nozoom_info)

stopifnot(nrow(nozoom_info) == 0L)
```

## Statistics validation

`stats_bwg()` supports `mean`, `stdev`, `max`, `min`, `coverage`, and `sum`.

### 10. Exact full-resolution statistics

```{r}
stats_full <- stats_bwg(
  roundtrip_lazy,
  chrom = "chr1",
  start = 1,
  end = 10000,
  n_bins = 10,
  stat = "mean",
  use_zoom = FALSE
)

print(stats_full)

stopifnot(
  nrow(stats_full) == 10L,
  all(stats_full$source == "full")
)
```

### 11. Zoom-accelerated statistics

```{r}
stats_zoom <- stats_bwg(
  roundtrip_lazy,
  chrom = "chr1",
  start = 1,
  end = 10000,
  n_bins = 10,
  stat = "mean",
  use_zoom = TRUE
)

print(stats_zoom)

stopifnot(
  nrow(stats_zoom) == 10L,
  all(stats_zoom$source %in% c("full", "zoom"))
)
```

For exact scientific calculations, use `use_zoom = FALSE`. Partial overlap with
a stored zoom window is approximate because only summary values are available at
that resolution.

## WIG and bedGraph round-trip

### 12. bedGraph

```{r}
bedgraph_outdir <- file.path(test_root, "bedgraph")

written_bedgraph <- write_bwg(
  track_bedgraph,
  outdir = bedgraph_outdir,
  format = "bedgraph",
  overwrite = TRUE
)

bedgraph_roundtrip <- read_bwg(
  written_bedgraph$file,
  format = "bedgraph",
  sample_names = "example",
  mode = "memory"
)

assert_signal_equal(
  track_bedgraph$data,
  bedgraph_roundtrip$data
)
```

### 13. WIG

```{r}
wig_outdir <- file.path(test_root, "wig")

written_wig <- write_bwg(
  track_bedgraph,
  outdir = wig_outdir,
  format = "wig",
  overwrite = TRUE
)

wig_roundtrip <- read_bwg(
  written_wig$file,
  format = "wig",
  sample_names = "example",
  mode = "memory"
)

assert_signal_equal(
  track_bedgraph$data,
  wig_roundtrip$data
)
```


## Multi-region retrieval

`retrieve_bwg()` is the only genomic extraction interface. It supports both the
original single-region API and a `regions` table for multi-region retrieval.
Extraction never writes files and does not expose writer-specific controls such
as `format`, `compress`, `zoom`, or `max_zoom_levels`.

Overlapping or directly adjacent query regions are merged before retrieval.
Signal intervals crossing a query boundary are clipped while keeping their
original genomic coordinate system.

### 14. Retrieve multiple regions as signal data

```{r}
retrieve_regions <- data.frame(
  chrom = c("chr1", "chr1", "chr2"),
  start = c(510L, 561L, 450L),
  end = c(560L, 600L, 900L)
)

retrieved_data <- retrieve_bwg(
  roundtrip_lazy,
  regions = retrieve_regions
)

print(retrieved_data)

stopifnot(
  nrow(retrieved_data) > 0L,
  all(retrieved_data$chrom %in% c("chr1", "chr2"))
)
```

The two adjacent `chr1` queries are normalized to the genomic union
`chr1:510-600`, preventing duplicate signal records.

### 15. Retrieve a write-ready BwgTrack

Use `result = "track"` when the extracted signal should remain inside the
standard `BwgTrack` object contract:

```{r}
retrieved_track <- retrieve_bwg(
  roundtrip_lazy,
  regions = retrieve_regions,
  result = "track"
)

print(retrieved_track)
print(retrieved_track$meta$retrieval_regions)

stopifnot(
  is_bwg_track(retrieved_track),
  identical(retrieved_track$meta$mode, "memory"),
  isTRUE(retrieved_track$meta$retrieved),
  nrow(retrieved_track$meta$retrieval_regions) == 2L
)
```

The default `result = "data"` is retained for compatibility with existing
GeneTrackR code that expects the signal data.table directly.

### 16. Persist retrieved data only with write_bwg()

File output is deliberately separated from retrieval:

```{r}
retrieved_bigwig_dir <- file.path(
  test_root,
  "retrieved_bigwig"
)

retrieved_bigwig <- write_bwg(
  retrieved_track,
  outdir = retrieved_bigwig_dir,
  format = "bigwig",
  overwrite = TRUE
)

stopifnot(
  nrow(retrieved_bigwig) == 1L,
  file.exists(retrieved_bigwig$file)
)
```

Read the new file back and compare it with the retrieved in-memory track:

```{r}
retrieved_roundtrip <- read_bwg(
  retrieved_bigwig$file,
  format = "bigwig",
  sample_names = "example",
  mode = "memory"
)

assert_signal_equal(
  retrieved_track$data,
  retrieved_roundtrip$data
)
```

This separation keeps responsibilities explicit:

```{text}
read_bwg()      open/load tracks
retrieve_bwg()  extract genomic regions
stats_bwg()     calculate interval statistics
write_bwg()     persist or convert tracks
```

## Coordinate contract

A bedGraph record:

```{text}
chr1    499    524    9.1000
```

is 0-based half-open and becomes:

```{text}
chr1    500    524    9.1000
```

in the public 1-based closed in-memory representation. The reverse conversion is
applied when writing BigWig or bedGraph.

## Package regression tests

Run after the manual checks:

```{r}
devtools::document()
devtools::test()
```

Then run:

```{r}
devtools::check()
```

Expected test status:

```{text}
FAIL 0
WARN 0
SKIP 0
```

Any reader, writer, zoom, statistics, object-contract, or text-format failure
should be treated as a regression before continuing development.

## Development status

- **0.1.x**: native reader and stable `BwgTrack` contract — implemented.
- **0.2.x**: native BigWig/WIG/bedGraph writer — implemented.
- **0.3.x**: BigWig zoom levels and interval statistics — implemented.
- **0.4.x**: unified indexed single-/multi-region retrieval — current stage.
- **0.5.x**: multi-track merge and explicit arithmetic collapse.
- **0.6.x**: GeneTrackR backend integration.
- **1.0.0**: stable API and cross-platform release.

See `DEVELOPMENT.md` for detailed acceptance criteria.
