---
title: "bwTools"
format: gfm
execute:
  echo: true
  eval: false
---

# bwTools

[![R-CMD-check](https://github.com/Renscq/bwTools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Renscq/bwTools/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/Renscq/bwTools/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/Renscq/bwTools/actions/workflows/pkgdown.yaml)

Current release: **0.8.7**

`bwTools` provides native-R tools for BigWig, WIG, and bedGraph genomic signal
tracks. It supports BigWig binary reading and writing, indexed regional
retrieval, exact and zoom-aware statistics, multi-track merging, format
conversion, and performance diagnostics without requiring Rcpp, bundled C/C++
code, or UCSC executables at runtime.

## Key features

- Native-R BigWig reader and writer with indexed random access.
- Native BigWig zoom summaries for fast large-region statistics.
- WIG and bedGraph reading and writing, including gzip-compressed text output.
- Standardized `BwgTrack` schema v2 for single- and multi-sample workflows.
- Exact strand-aware retrieval for in-memory tracks.
- Conservative lossless merging and explicit mean/sum aggregation.
- Stable `bwTools_error` conditions for downstream packages.
- Optional external compatibility checks for pyBigWig, rtracklayer, and UCSC
  utilities without adding runtime dependencies.

## Installation

Install the development release from GitHub:

```{r}
install.packages("pak")
pak::pak("Renscq/bwTools")
```

Package website: <https://renscq.github.io/bwTools/>

For package development and local validation:

```{r}
devtools::document()
devtools::test()
devtools::check()
```

## Quick start

The package ships a small BigWig example:

```{r}
library(bwTools)

bw_file <- system.file(
  "extdata",
  "bwtools_example.bigwig",
  package = "bwTools",
  mustWork = TRUE
)
```

Read it lazily:

```{r}
bw <- read_bwg(
  bw_file,
  sample_ids = "example"
)

summary_bwg(bw)
samples_bwg(bw)
seqinfo_bwg(bw)
```

Retrieve a genomic interval using public **1-based closed** coordinates:

```{r}
signal <- retrieve_bwg(
  bw,
  chrom = "chr1",
  start = 500L,
  end = 1600L
)
```

Calculate exact statistics:

```{r}
stats <- stats_bwg(
  bw,
  chrom = "chr1",
  start = 1L,
  end = 10000L,
  n_bins = 10L,
  stat = "mean",
  use_zoom = FALSE
)
```

Create a writable regional track and persist it:

```{r}
region_track <- retrieve_bwg(
  bw,
  chrom = "chr1",
  start = 1L,
  end = 10000L,
  result = "track"
)

manifest <- write_bwg(
  region_track,
  outdir = "bwtools-output",
  format = "bigwig",
  overwrite = TRUE
)
```

The standard data flow is:

```{text}
file
  ↓
read_bwg()
  ↓
BwgTrack
  ↓
retrieve_bwg() / merge_bwg() / stats_bwg()
  ↓
write_bwg()
```

## Coordinates and object contract

Public in-memory coordinates are always **1-based closed**. BigWig and bedGraph
coordinates on disk are **0-based half-open**; bwTools handles the conversion at
the file boundary.

A `BwgTrack` contains:

```{text}
BwgTrack
├── samples
├── data
├── seqinfo
└── meta
```

Use `bwg_track()` to construct a track explicitly, `validate_bwg()` to validate
the public contract, and `is_bwg_track()` for a logical contract check. Access
metadata through `samples_bwg()`, `seqinfo_bwg()`, `metadata_bwg()`, and
`zoominfo_bwg()` rather than relying on private implementation details.

## Multiple samples and strands

`read_bwg()` accepts multiple files with explicit sample IDs and strand labels:

```{r}
bedgraph_file <- system.file(
  "extdata",
  "bwtools_example.bedGraph",
  package = "bwTools",
  mustWork = TRUE
)

mixed <- read_bwg(
  c(bw_file, bedgraph_file),
  sample_ids = c("rna_plus", "rna_minus"),
  strand = c("+", "-")
)
```

BigWig-only multi-file input remains lazy. Mixing BigWig with WIG or bedGraph
uses memory mode. In-memory mixed-strand tracks are filtered from row-level
`data$strand`; lazy BigWig tracks use file-level strand metadata.

## Documentation

The user workflow is split into focused articles:

| Article | Scope |
| --- | --- |
| [Getting started](https://renscq.github.io/bwTools/articles/getting-started.html) | End-to-end workflow and `BwgTrack` basics |
| [Reading and writing](https://renscq.github.io/bwTools/articles/reading-writing.html) | BigWig/WIG/bedGraph I/O and manifests |
| [Retrieval and statistics](https://renscq.github.io/bwTools/articles/retrieve-stats.html) | Regions, boundaries, exact and zoom statistics |
| [Merge and multi-sample workflows](https://renscq.github.io/bwTools/articles/merge-multisample.html) | Sample identity, strands, aggregation |
| [File formats and coordinates](https://renscq.github.io/bwTools/articles/file-formats.html) | Format detection and coordinate conventions |
| [Performance and compatibility](https://renscq.github.io/bwTools/articles/performance-compatibility.html) | Benchmarking and external validation |

## Public API

| Function | Responsibility |
| --- | --- |
| `detect_bwg_format()` | Detect BigWig, WIG, or bedGraph |
| `read_bwg()` | Read one or more signal files |
| `bwg_track()` | Construct a standardized track |
| `validate_bwg()` | Validate the public object contract |
| `is_bwg_track()` | Test the object contract |
| `samples_bwg()` | Access sample metadata |
| `seqinfo_bwg()` | Access chromosome metadata |
| `metadata_bwg()` | Access object metadata |
| `zoominfo_bwg()` | Access stored BigWig zoom metadata |
| `retrieve_bwg()` | Retrieve one or more genomic regions |
| `merge_bwg()` | Merge or aggregate tracks |
| `stats_bwg()` | Calculate interval statistics |
| `summary_bwg()` | Summarize a track |
| `write_bwg()` | Write BigWig, WIG, or bedGraph |
| `benchmark_bwg()` | Profile large BigWig workflows |

The 15-function public API and BwgTrack schema v2 are regression-tested.

## Errors

Intentional user-facing errors inherit from `bwTools_error`, allowing downstream
packages to catch package errors without matching English text:

```{r}
tryCatch(
  retrieve_bwg(bw, "chr1", 0L, 100L),
  bwTools_error = function(e) conditionMessage(e)
)
```

## Performance diagnostics

`benchmark_bwg()` is a development and profiling utility rather than a routine
analysis step. See [Performance and compatibility](https://renscq.github.io/bwTools/articles/performance-compatibility.html)
for reader, statistics, and writer profiling examples.

## External compatibility validation

Optional bidirectional BigWig checks are installed under
`inst/compatibility/`. They can use pyBigWig/libBigWig, rtracklayer, and UCSC
utilities when those tools are available, but none is a bwTools runtime
dependency.
