---
title: "File formats and coordinate conventions"
description: "Understand format detection, coordinate conversion, chromosome names, sequence metadata, and zoom metadata."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{File formats and coordinate conventions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Supported formats

`detect_bwg_format()` returns one of `bigwig`, `wig`, or `bedgraph`.
BigWig is detected from its four-byte binary magic signature before text
parsing, so a valid BigWig does not depend on its file extension.

```{r}
library(bwTools)

bw_file <- system.file("extdata", "bwtools_example.bigwig", package = "bwTools")
bg_file <- system.file("extdata", "bwtools_example.bedGraph", package = "bwTools")

detect_bwg_format(bw_file)
detect_bwg_format(bg_file)
```

# Coordinate systems

bwTools exposes one public in-memory coordinate convention:


    1-based closed


A public interval `start = 1, end = 10` contains 10 bases.

On disk:


    BigWig   0-based half-open
    bedGraph 0-based half-open
    WIG      format-specific text representation


The conversion is performed only at the read/write boundary. Do not manually
subtract or add one when calling `retrieve_bwg()` or `stats_bwg()`.

# Chromosome names

Chromosome identifiers are preserved exactly. bwTools does not rewrite `1` to
`chr1`, normalize case, or map mitochondrial aliases. Therefore `1`, `chr1`,
`Chr01`, `MT`, and `scaffold_001` are distinct names.

# Sequence metadata

`seqinfo_bwg()` returns:


    sample_id  chrom  length


BigWig stores chromosome lengths, so lazy BigWig tracks normally have complete
sequence metadata. WIG and bedGraph do not encode complete chromosome sizes, so
inferred lengths may be `NA` unless explicit metadata are supplied through
`bwg_track()` or `write_bwg(..., chrom_sizes = ...)`.

# BigWig zoom metadata

`zoominfo_bwg()` returns stored reduction levels for BigWig-backed samples:

```{r}
bw <- read_bwg(bw_file, sample_ids = "example")
zoominfo_bwg(bw)
```

WIG, bedGraph, and memory-only tracks have no stored BigWig zoom index and
return a typed zero-row zoom table.

# Format conversion

All conversion follows the same object path:


    source file -> read_bwg() -> BwgTrack -> write_bwg() -> target file


This keeps format-specific coordinate conversion out of retrieval, statistics,
and merging code.
