---
title: "Getting started with bwTools"
description: "Read, inspect, retrieve, summarize, merge, and write signal tracks through the standard BwgTrack workflow."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bwTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Overview

`bwTools` uses one stable data flow for BigWig, WIG, and bedGraph signal tracks:


    file -> read_bwg() -> BwgTrack -> analysis -> write_bwg()


Public coordinates are always **1-based closed**.

# Load the bundled example

```{r}
library(bwTools)

bw_file <- system.file(
  "extdata",
  "bwtools_example.bigwig",
  package = "bwTools",
  mustWork = TRUE
)

bw <- read_bwg(bw_file, sample_ids = "example")
```

A BigWig-only input uses lazy indexed access by default. Inspect the track with
`summary_bwg()`, `samples_bwg()`, `seqinfo_bwg()`, `metadata_bwg()`, and
`zoominfo_bwg()`:

```{r}
summary_bwg(bw)
samples_bwg(bw)
seqinfo_bwg(bw)
metadata_bwg(bw)
zoominfo_bwg(bw)
```

Use `is_bwg_track()` for a logical contract check and `validate_bwg()` when an
invalid object should raise an error:

```{r}
is_bwg_track(bw)
validate_bwg(bw)
```

# Retrieve a region

```{r}
region <- retrieve_bwg(
  bw,
  chrom = "chr1",
  start = 500L,
  end = 1600L
)
```

Request `result = "track"` when the result will be passed to another bwTools
operation:

```{r}
region_track <- retrieve_bwg(
  bw,
  chrom = "chr1",
  start = 500L,
  end = 1600L,
  result = "track"
)
```

# Calculate statistics

```{r}
stats_bwg(
  bw,
  chrom = "chr1",
  start = 1L,
  end = 10000L,
  n_bins = 20L,
  stat = "mean",
  use_zoom = FALSE
)
```

Use `use_zoom = TRUE` for fast large-region summaries when approximate boundary
handling from stored BigWig zoom windows is acceptable.

# Merge tracks

`merge_bwg()` conservatively preserves sample identity with `method = "auto"`:

```{r}
a <- retrieve_bwg(bw, "chr1", 500L, 900L, result = "track")
b <- retrieve_bwg(bw, "chr1", 1200L, 1600L, result = "track")
merged <- merge_bwg(a, b)
```

Use `method = "mean"` or `method = "sum"` only when arithmetic aggregation is
intended.

# Write a result

```{r}
manifest <- write_bwg(
  region_track,
  outdir = "bwtools-output",
  format = "bigwig",
  overwrite = TRUE
)
```

`write_bwg()` is the only normal signal-file persistence function. The returned
manifest has one row per selected written sample.

# Constructing a BwgTrack directly

Use `bwg_track()` when signal data already exist in R:

```{r}
x <- bwg_track(
  samples = data.frame(sample_id = "sampleA"),
  data = data.frame(
    sample_id = "sampleA",
    chrom = "chr1",
    start = 1L,
    end = 10L,
    value = 2,
    strand = "+"
  ),
  seqinfo = data.frame(chrom = "chr1", length = 1000L)
)
```

The in-memory signal schema is `sample_id`, `chrom`, `start`, `end`, `value`,
and `strand`.
