---
title: "Merging and multi-sample workflows"
description: "Work with multiple samples and strands and choose conservative or arithmetic merge semantics."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merging and multi-sample workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Multi-sample input

`read_bwg()` preserves sample identity, source format, and file-level strand
metadata.

```{r}
library(bwTools)

bw_file <- system.file("extdata", "bwtools_example.bigwig", package = "bwTools")
bg_file <- system.file("extdata", "bwtools_example.bedGraph", package = "bwTools")

x <- read_bwg(
  c(bw_file, bg_file),
  sample_ids = c("sampleA", "sampleB"),
  strand = c("+", "-")
)

samples_bwg(x)
seqinfo_bwg(x)
```

BigWig-only multi-file input stays lazy. Mixed BigWig/WIG/bedGraph input uses
memory mode.

# Conservative automatic merge

`merge_bwg(..., method = "auto")` is lossless and non-arithmetic:

```{r}
a <- retrieve_bwg(x, "chr1", 1L, 1000L, sample_ids = "sampleA", result = "track")
b <- retrieve_bwg(x, "chr1", 2000L, 3000L, sample_ids = "sampleA", result = "track")

combined <- merge_bwg(a, b)
```

Different sample IDs remain separate. Disjoint signal for the same sample is
appended. Exact duplicates and equal-valued overlaps are normalized. Conflicting
same-sample overlaps raise `bwTools_error` rather than silently choosing an
arithmetic rule.

# Mean and sum aggregation

Use arithmetic methods explicitly:

```{r}
mean_track <- merge_bwg(
  x,
  method = "mean",
  missing = "ignore"
)

sum_track <- merge_bwg(
  x,
  method = "sum"
)
```

For means, `missing = "ignore"` averages only samples covering an atomic segment;
`missing = "zero"` treats uncovered members as zero.

# Grouped aggregation

```{r}
groups <- data.frame(
  sample_id = c("rep1", "rep2", "rep3", "rep4"),
  group_id = c("control", "control", "treated", "treated")
)

# merge_bwg(track, method = "mean", groups = groups)
```

`groups` is used only with `method = "mean"` or `method = "sum"`.

# Mixed strands after merging

A memory-mode sample can legitimately contain both `+` and `-` signal rows.
Its sample-level metadata becomes `strand = "*"`, but retrieval remains exact
because row-level signal strand is used:

```{r}
plus_track <- bwg_track(
  samples = data.frame(sample_id = "sampleA", strand = "+"),
  data = data.frame(
    sample_id = "sampleA", chrom = "chr1",
    start = 1L, end = 10L, value = 1, strand = "+"
  )
)
minus_track <- bwg_track(
  samples = data.frame(sample_id = "sampleA", strand = "-"),
  data = data.frame(
    sample_id = "sampleA", chrom = "chr1",
    start = 20L, end = 30L, value = 1, strand = "-"
  )
)
mixed_strand <- merge_bwg(plus_track, minus_track)

plus <- retrieve_bwg(mixed_strand, "chr1", 1L, 50L, strand = "+")
minus <- retrieve_bwg(mixed_strand, "chr1", 1L, 50L, strand = "-")
```

Use `strand = "ignore"` in `merge_bwg()` only when combining strands is an
explicit analytical decision.
