---
title: "Reading and writing signal tracks"
description: "Read BigWig, WIG, and bedGraph files and write standardized signal outputs and manifests."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and writing signal tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Reading files

`read_bwg()` accepts BigWig, WIG, and bedGraph inputs. `format = "auto"` calls
`detect_bwg_format()` independently for each file.

```{r}
library(bwTools)

bw_file <- system.file("extdata", "bwtools_example.bigwig", package = "bwTools")
bg_file <- system.file("extdata", "bwtools_example.bedGraph", package = "bwTools")

detect_bwg_format(bw_file)
detect_bwg_format(bg_file)

bw <- read_bwg(bw_file, sample_ids = "bw")
bg <- read_bwg(bg_file, sample_ids = "bg")
```

BigWig-only inputs use lazy mode by default. WIG and bedGraph are materialized
in memory. Mixed binary/text input also uses memory mode:

```{r}
mixed <- read_bwg(
  c(bw_file, bg_file),
  sample_ids = c("binary", "text")
)
summary_bwg(mixed)
```

# Sample and strand metadata

Use unique `sample_ids` and optional file-level strand labels:

```{r}
tracks <- read_bwg(
  c(bw_file, bg_file),
  sample_ids = c("plus", "minus"),
  strand = c("+", "-")
)

samples_bwg(tracks)
```

# Writing files

`write_bwg()` writes one file per selected sample and returns a manifest with
`sample_id`, `file`, and `format`.

```{r}
region <- retrieve_bwg(
  bw,
  chrom = "chr1",
  start = 1L,
  end = 5000L,
  result = "track"
)

manifest <- write_bwg(
  region,
  outdir = "bigwig-out",
  format = "bigwig",
  overwrite = TRUE
)
```

Native BigWig output contains zoom levels by default. Disable them only when a
full-resolution-only file is required:

```{r}
write_bwg(
  region,
  outdir = "bigwig-nozoom",
  format = "bigwig",
  overwrite = TRUE,
  zoom = FALSE
)
```

# WIG and bedGraph output

```{r}
write_bwg(
  region,
  outdir = "wig-out",
  format = "wig",
  overwrite = TRUE
)

write_bwg(
  region,
  outdir = "bedgraph-out",
  format = "bedgraph",
  overwrite = TRUE,
  compress = TRUE
)
```

`compress = TRUE` applies to WIG/bedGraph text output. BigWig block compression
is part of the BigWig format and is handled internally.

# Chromosome sizes

BigWig output requires chromosome lengths. Complete lengths are taken from
`seqinfo_bwg()` when available, or supplied explicitly with `chrom_sizes`:

```{r}
chrom_sizes <- data.frame(
  chrom = "chr1",
  length = 100000L
)

write_bwg(
  region,
  outdir = "bigwig-explicit-sizes",
  format = "bigwig",
  chrom_sizes = chrom_sizes,
  overwrite = TRUE
)
```

For multi-sample in-memory tracks, every selected sample must contain signal.
`write_bwg()` checks this before creating any signal file so partial manifests
are not produced silently.
