---
title: "Regional retrieval and statistics"
description: "Use indexed regional retrieval, chromosome-boundary rules, and exact or zoom-aware interval statistics."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regional retrieval and statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

# Regional retrieval

`retrieve_bwg()` accepts either one interval or a table of intervals. Public
coordinates are **1-based closed**.

```{r}
library(bwTools)

bw_file <- system.file("extdata", "bwtools_example.bigwig", package = "bwTools")
bw <- read_bwg(bw_file, sample_ids = "example")

one <- retrieve_bwg(bw, "chr1", 500L, 1600L)
```

For several regions:

```{r}
regions <- data.frame(
  chrom = c("chr1", "chr1", "chr2"),
  start = c(510L, 561L, 450L),
  end = c(560L, 600L, 900L)
)

many <- retrieve_bwg(bw, regions = regions)
```

Overlapping or directly adjacent query regions are normalized before retrieval.
Signal intervals crossing a boundary are clipped but are never rebased.

# Boundary behavior

When complete chromosome lengths are known:


    unknown chromosome             -> error
    end beyond chromosome length   -> clipped
    start beyond chromosome length -> typed empty result


WIG and bedGraph do not encode chromosome lengths. Their inferred `seqinfo`
therefore commonly has `length = NA`; unknown names in such incomplete metadata
are treated as no signal rather than as an invalid reference chromosome.

# Strand filtering

```{r}
plus <- retrieve_bwg(
  bw,
  chrom = "chr1",
  start = 1L,
  end = 5000L,
  strand = "+"
)
```

Memory-mode tracks filter row-level `strand` values. Lazy BigWig tracks use
file-level strand metadata.

# Exact statistics

`stats_bwg()` supports `mean`, `stdev`, `max`, `min`, `coverage`, and `sum`:

```{r}
exact <- stats_bwg(
  bw,
  chrom = "chr1",
  start = 1L,
  end = 10000L,
  n_bins = 100L,
  stat = "mean",
  use_zoom = FALSE
)
```

Exact lazy-BigWig statistics stream overlapping full-resolution blocks into bin
accumulators instead of first materializing the complete signal table.

# Zoom-aware statistics

```{r}
fast <- stats_bwg(
  bw,
  chrom = "chr1",
  start = 1L,
  end = 1000000L,
  n_bins = 1000L,
  stat = "mean",
  use_zoom = TRUE
)
```

Inspect `source` and `resolution` in the result. A small high-resolution query
may still use `source = "full"` when no stored zoom level is appropriate.
Use `use_zoom = FALSE` when exact query-boundary values are required.
